---
title: "Case Study: Part 3: Decision Trees and Gradient Boosting"
author: "Your NAME"
bibliography: ../assets/references_intro.bib
csl: ../assets/apa.csl
link-citations: true
format:
  html:
    theme: 
       light: cerulean
    # theme: darkly
    # highlight: espresso
    code-copy: true
    code-fold: true
    df-print: paged
    include-in-header: ../assets/mathjax.html
    number-sections: true
    toc: true
    toc_depth: 3
    toc_float: yes
    toc-location: left
    fontsize: 10pt
    mainfont: "Helvetica Neue"
  pdf:
    number-sections: true
    toc: true
    toc_depth: 2
    toc_float: yes
execute: 
  #cache: true
  warning: false
fontsize: 11pt
---

# The Farm holding / Desbois Case Study

## Context

The article from @desbois2008 (available [here](https://csbigs.fr/index.php/csbigs/article/view/351.html) together with a sample data set) presents a complete case study around the detection of financial risks for **farm holdings** in France.

Among other methods, Linear Discriminant Analysis and Logistic Regression (plus a stepwise variable selection procedure) are used. ROC curves are provided to compare the methods.

We will reproduce and extend the Desbois case study throughout this course as a **running example** ("fil rouge"). At each stage, we apply the concepts introduced in the preceding handout.

::: callout-tip
## What you'll do in this Part 3

Following the concepts from [Handout 3: Decision Trees](../handouts/3_decision_trees.qmd):

1.  Fit various decision trees on Farm holding data using CART algorithm. Understand their main parameters, what prunning means. Assess various trees using resampling methods
2.  Experiment with Gradient Boosting algorithms. Understand their main parameters, losses used


:::

## Data

### Loading

Loading Desbois data and first glimpse at it:

```{r}
#| echo: true
#| code-fold: show
#| warning: false

library(tidyverse)
# package used to import spss file
library(foreign)

don_desbois <- read.spss("../data/presentation_data/Agriculture Farm Lending/desbois.sav", to.data.frame = TRUE) %>% as_tibble()
glimpse(don_desbois)
```

Replacing for convenience `DIFF` target by factor `Y` with `O` (healthy) or `1` (failing):

```{r}
#| echo: true
#| code-fold: show
#| warning: false

don_desbois <- don_desbois %>%
    mutate(Y = as.factor(if_else(DIFF=='healthy', 0, 1)), DIFF = NULL,
           .before = everything())

```

Your turn to play.

## CART Algorithm

Using the Agriculture Farm Lending data set:

-   `CART/rpart` build and plot a decision tree of you choice using `rpart`.

    You can use default parameters, and this way learning them. (see for example 2. Building the tree [here](https://cran.r-project.org/web/packages/rpart/vignettes/longintro.pdf)

Try to display class probabilities and number of observations per node/leaf (for that explore functions `rpart.plot` or `pdp` from package `rpart.plot` see [vignette](http://www.milbo.org/doc/prp.pdf)).

```{r}
#| echo: true
#| code-fold: show
#| warning: false
# YOUR CODE HERE
```

-   Customize the tree plot to show both class probabilities and number of observations per node/leaf.

```{r}
#| echo: true
#| code-fold: show
#| warning: false
# YOUR CODE HERE
```

-   Playing with `rpart` parameters (in particular those inside `rpart.control`), fit, plot, then compare a "large/deep" decision tree and a "smaller/shallower" tree in terms of ROC/AUC/prediction on the holdout set.

```{r}
#| echo: true
#| code-fold: show
#| warning: false
# YOUR CODE HERE
```

-   Describe in simple terms the output of the `printcp` function (see for example 4. Pruning the tree [here](https://cran.r-project.org/web/packages/rpart/vignettes/longintro.pdf). You don't have to understand deeply the pruning process but understand what is at stake in the process.

-   Choose an arbitrary terminal number of leafs and prune the tree using the `prune` function.

```{r}
#| echo: true
#| code-fold: show
#| warning: false
# YOUR CODE HERE
```

-   Select the optimal `cp` parameter for your tree (you can use a plot) and compare to the "large" and "small" models in terms of AUC.

```{r}
#| echo: true
#| code-fold: show
#| warning: false
# YOUR CODE HERE
```

- Compare now the deep/pruned/shallow trees in terms of AUC using a resampling procedure (Hold-out or K-fold CV).

```{r}
#| echo: true
#| code-fold: show
#| warning: false
# YOUR CODE HERE
```

## Gradient boosting

Using a gradient boosting package of your choice (`gbm`, `xgboost`), play with number of boosting iterations, weak learner complexity (decision tree depth, min number of observations), learning rate. Compare also logistic loss with exponential loss (adaboost) as in the slides/handouts.

Note that the data at hand is quite small (training set is around 1000 observations).

You can take inspiration from this graph: [here](https://scikit-learn.org/1.5/auto_examples/ensemble/plot_gradient_boosting_regularization.html))

Example usage of `gbm` (more [here](https://cran.r-project.org/web/packages/gbm/vignettes/gbm.pdf), see for example the Figure 3 showing Out-of-sample predictive performance by number of iterations and shrinkage):

```{r}
#| echo: true
#| code-fold: show
#| warning: false
# YOUR CODE HERE
# library(gbm)
# gbm_deg <- gbm(Y~.,
#            data = YOUR_DATA,
#            n.trees = N, # number of boosting iterations
#            distribution = "bernoulli", # loss minimized
#            interaction.depth = 1, 
#            /!\ interaction.depth (~#{terminal nodes}+1) <> rpart maxdepth
#                n.minobsinnode = 1, # comparable to rpart minbucket
#            shrinkage = 1, # learning rate
#            bag.fraction = 1) # set at 1 to implement pure boosting 
#            # (otherwise stochastic boosting, ie mix of boosting and bagging)
```

Example usage of `xgboost` (more [here]() or [here](https://xgboost.readthedocs.io/en/stable/R-package/xgboostPresentation.html) for the `R` package):

```{r}
#| echo: true
#| code-fold: show
#| warning: false
# YOUR CODE HERE
#library(xgboost)
# xgb_deg <- xgboost(data = as.matrix(YOUR_PREDICTORS),
#            label = as.vector(YOUR_TARGET),
#            max.depth = 1, # comparable to rpart maxdepth
#            eta = 1, # learning rate
#            nthread = 1,
#            nrounds = num_tree, # number of boosting iterations
#             min_child_weight = 0, # similar but not comparable to rpart minbucket
#             objective = "binary:logistic", # loss minimized
#             lambda = 0, # set at 0 to avoid L2 penalization
#             tree_method = "exact") 
```